The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-11-30 article_final/main.tex
Added charts for Medium GPT

\begin{abstract}
	In this article, we explore the use of various matrix norms for optimizing functions of weight matrices, a crucial problem in training large language models. Moving beyond the spectral norm underlying the Muon update, we leverage duals of the Ky Fan $k$-norms to introduce a family of Muon-like algorithms we name \emph{Fanions}, which are closely related to Dion. By working with duals of convex combinations of the Ky Fan $k$-norms with either the Frobenius norm or the $l_\infty$ norm, we construct the families of \emph{F-Fanions} and \emph{S-Fanions}, respectively. Their most prominent members are \emph{F-Muon} and \emph{S-Muon}. We complement our theoretical analysis with an extensive empirical study of these algorithms across a wide range of tasks and settings, demonstrating that F-Muon and S-Muon consistently match Muon's performance, while outperforming vanilla Muon on synthetic linear least squares problems.
\end{abstract}

@@

Most notably, our experiments on real-world tasks demonstrate that the choice of underlying matrix norm is remarkably flexible. On CIFAR-10 airbench, properly-tuned F-Muon and S-Muon achieve $94.02 \pm 0.13\%$ and $94.03 \pm 0.13\%$ accuracy, matching Muon's $94.01 \pm 0.13\%$ performance. On NanoGPT pre-training, F-Muon achieves 3.281 cross-entropy loss, while fully-tuned Muon achieves 3.279. Finally, S-Muon matches Muon on fine-tuning NanoGPT on Tiny Stories, while F-Muon is far more resistant to learning rate choice than Muon. These results show that Muon-like algorithms can maintain competitive performance even when the underlying norm constraint is significantly modified, providing an affirmative answer to the central question posed above. Moreover, the tools from \sectionref{sec:conic_combo} give researchers considerable flexibility in designing algorithms that need not be direct modifications of Muon.

@@

Unlike Schatten norms, which satisfy $\norm{\mM^t}_{S_p}^\dagger = \norm{\mM^t}_{S_q}$ (for $p^{-1} + q^{-1} = 1$) and are thus closed under dualization, Ky Fan norms are generally not. Only two exceptional cases exist: the dual of the Ky Fan $1$-norm $\norm{\cdot}_{\text{KF-}1}$ (the spectral norm) is the Ky Fan $\min(m,n)$-norm $\norm{\cdot}_{\text{KF-}\min(m,n)}$ (the nuclear norm), and vice versa. Thus working with duals of Ky Fan norms can open up new possibilities:

@@

By \corollaryref{corollary:lin_comb_alg}, this defines an LMO-algorithm with norm $\norm{\cdot}_{\mathrm{C-KF-k}}^\dagger$, where
\begin{equation}\label{eq:ckfk_norm}
	\norm{\cdot}_{\mathrm{C-KF-k}} = \alpha \normkfk{\cdot} + \frac{1-\alpha}{\eta}\normc{\cdot}^\dagger\,.sec:lmo_for_nn
\end{equation}

@@

\subsection{Randomized Linear Least Squares}
\label{subsec:lls_exps}

@@

\begin{figure}[htbp]
\floatconts
  {fig:cifar_ball}
  {\caption{Visualization of the LMO balls for Muon and F-Muon.}}
  {\includegraphics[width=0.6\linewidth]{figs/norms/fstardual_cifar.pdf}}
\end{figure}

@@

We evaluate F-Muon and S-Muon with $\alpha=0.5$ against Muon, SignSGD, and NSGD on the NanoGPT speedrun benchmark \citep{modded_nanogpt_2024}. The optimal hyperparameters are shown in the legend of \figureref{fig:gpt_loss}, with \texttt{sign\_sgd\_coeff} values of 0.0003/0.07 for S-Muon. After 1750 training steps, F-Muon achieves a cross-entropy loss of 3.281, S-Muon achieves 3.287, and Muon achieves 3.279, falling below the target threshold of 3.280. As illustrated in \figureref{fig:gpt_loss}, these differences are negligible. Notably, F-Muon represents a convex combination of Muon and NSGD, despite NSGD performing poorly in isolation. The same observation holds for S-Muon.

\begin{figure}[htbp]
\floatconts
  {fig:gpt_loss}
  {\caption{The validation loss for NanoGPT}}
  {\includegraphics[width=0.7\linewidth]{figs/nanogpt/val_loss_vs_step.pdf}}
\end{figure}

As observed in the CIFAR airbench experiments, setting $\alpha=-0.1$ for F-Muon (corresponding to the no-ball configuration) yields a loss of 3.2818, representing only a marginal difference from the baseline.

Notably, F-Muon with the NSGD component not scaled by Muon's rule yields a loss of 3.288 (an increase of 0.007), while S-Muon scaled with Muon's rule achieves a loss of 3.292 (an increase of 0.005).

@@

@@

The updates of LMO optimizers exhibit striking similarities between vector $l_p$ norms and matrix Schatten $S_p$ norms, as illustrated in \tableref{tbl:mat_vs_vec_lmo}. These parallels extend beyond the update formulas themselves to empirical performance characteristics. SignSGD closely resembles Adam, as noted in \citep{bernstein2024oldoptimizernewnorm}, and both Adam and Muon perform well when training large models \citep{zhao2025deconstructing, liu2025muon}. NSGD maintains identical updates in both vector and matrix cases. Greedy Coordinate Descent is rarely applied to high-dimensional problems, which provides perspective on why one-rank Neon underperforms in such settings.

@@

A common limitation of these analyses is their focus on convergence measured by the gradient norm. As we showed in our CIFAR experiments \cref{sec:cifar_underhood}, the gradient norm may decrease by only a factor of ten when the accuracy reaches 100\%.

@@

We generalized several successful algorithms, including Muon, Dion, and $\nu$SAM, to LMO-based algorithms using the family of norms dual to Ky Fan $k$-norms, yielding the Fanion family with low-rank updates. We further proposed the technique of regularizing these updates by combining them with NSGD or SignSGD, creating the F-Fanion and S-Fanion families through the $\norm{\cdot}_{\mathrm{F-KF-k}}$ and $\norm{\cdot}_{\mathrm{C^\dagger-KF-k}}$ norms. Our experiments demonstrate that F-Muon and S-Muon achieve competitive performance with Muon on CIFAR-10 airbench and NanoGPT tasks, confirming that the underlying norm constraint can be significantly modified without sacrificing effectiveness.

@@

\section{Author Contributions}

IO suggested using the nuclear norm in the \citep{bernstein2024oldoptimizernewnorm} framework. DM presented the problem at the MIPT optimization course. IK suggested using composite norms (though not the ones that induce F-Fanions and S-Fanions) and helped to draft and rewise the manuscript. NK suggested Lanczos algorithm as the most precise means to compute Fanions' updates, conducted experiments to prove this, and wrote \sectionref{sec:matrix-side}. AV conducted the finetuning of NanoGPT on Tiny Stories and wrote \sectionref{subsec:finetune}. All other work was done by AK: constructing Fanions, F-Fanions, S-Fanions, experimenting on the Linear Least Squares, CIFAR-airbench and NanoGPT pretrain, and writing the manuscript.

@@

\section{Proof of Lemma on Dual of Convex Combinations}\label{sec:proof_dual_conv_comb}

@@

\begin{figure}[htbp]
\floatconts
  {fig:app_lls}
  {\caption{More images for Linear least squares problem for a 500x500 matrix}}
  {%
    \subfigure[The spectral norm of the gradient]{\label{fig:lls_op_grad_norm}%
      \includegraphics[width=0.6\linewidth]{figs/simple_lls/gradient_spectral_norm_vs_iteration_long_run.pdf}}%
    \vspace{1em}
    \subfigure[The nuclear norm of the gradient]{\label{fig:lls_nuclear_grad_norm}%
      \includegraphics[width=0.6\linewidth]{figs/simple_lls/gradient_nuclear_norm_vs_iteration_long_run.pdf}}%
    \vspace{1em}
    \subfigure[The loss over time]{\label{fig:lls_loss_time}%
      \includegraphics[width=0.6\linewidth]{figs/simple_lls/loss_vs_time_long_run.pdf}}
  }
\end{figure}

@@

Most bounds for Muon and other LMO-algorithms are given for the norm of the gradient \citep{kovalev2025understanding, pethick2025training, riabinin2025gluon,kovalev2025noneuclideansgdstructuredoptimization}. Accordingly, it is natural to measure those norms on a real deep learning problem.

@@

We compare Muon, F-Muon, S-Muon, NSGD, SignSGD, Neon, and F-Neon, Fanion-5, and F-Fanion-5 on CIFAR-airbench. NSGD and SignSGD are not heavily tuned. The hyperparameters of Neon, F-Neon, Fanion-5, and F-Fanion-5 are taken from Muon and F-Muon. The parameters are in the table \tableref{tbl:cifar_lrs}. val\_accuracy is for the the weighted variant. However, to preserve the algorithms, we do not normalize the weights of the network each iteration.

>>>>

\begin{abstract}
	In this article, we explore the use of various matrix norms for optimizing functions of weight matrices, a crucial problem in training large language models. Moving beyond the spectral norm underlying the Muon update, we leverage duals of the Ky Fan $k$-norms to introduce a family of Muon-like algorithms we name \emph{Fanions}, which are closely related to Dion. By working with duals of convex combinations of the Ky Fan $k$-norms with either the Frobenius norm or the $l_\infty$ norm, we construct the families of \emph{F-Fanions} and \emph{S-Fanions}, respectively. Their most prominent members are \emph{F-Muon} and \emph{S-Muon}. We complement our theoretical analysis with an extensive empirical study of these algorithms across a wide range of tasks and settings, demonstrating that F-Muon and S-Muon consistently match Muon's performance, while outperforming vanilla Muon on a synthetic linear least squares problem.
\end{abstract}

@@

Most notably, our experiments on real-world tasks demonstrate that the choice of underlying matrix norm is remarkably flexible. On CIFAR-10 airbench, properly-tuned F-Muon and S-Muon achieve $94.02 \pm 0.13\%$ and $94.03 \pm 0.13\%$ accuracy, matching Muon's $94.01 \pm 0.13\%$ performance. On pre-training of NanoGPT F-Muon achieves 3.281 cross-entropy loss, while fully-tuned Muon achieves 3.279; of GPT-2 Medium - F-Muon 2.9215 and Muon 2.9198. Finally, S-Muon matches Muon on fine-tuning NanoGPT on Tiny Stories, while F-Muon is far more resistant to learning rate choice than Muon. These results show that Muon-like algorithms can maintain competitive performance even when the underlying norm constraint is significantly modified, providing an affirmative answer to the central question posed above. Moreover, the tools from \sectionref{sec:conic_combo} give researchers considerable flexibility in designing algorithms that need not be direct modifications of Muon.

@@

Unlike Schatten norms, which satisfy $\norm{\mM^t}_{S_p}^\dagger = \norm{\mM^t}_{S_q}$ (for $p^{-1} + q^{-1} = 1$) and are thus closed under dualization, Ky Fan norms are generally not. Only two exceptional cases exist: the dual to the Ky Fan $1$-norm $\norm{\cdot}_{\text{KF-}1}$ (the spectral norm) is the Ky Fan $\min(m,n)$-norm $\norm{\cdot}_{\text{KF-}\min(m,n)}$ (the nuclear norm), and vice versa. Thus working with duals of Ky Fan norms can open up new possibilities:

@@

By \corollaryref{corollary:lin_comb_alg}, this defines an LMO-algorithm with norm $\norm{\cdot}_{\mathrm{C-KF-k}}^\dagger$, where
\begin{equation}\label{eq:ckfk_norm}
	\norm{\cdot}_{\mathrm{C-KF-k}} = \alpha \normkfk{\cdot} + \frac{1-\alpha}{\eta}\normc{\cdot}^\dagger\,.
\end{equation}

@@

\subsection{Linear Least Squares}
\label{subsec:lls_exps}

@@

\begin{figure}[htbp]
\floatconts
  {fig:cifar_ball}
  {\caption{Visualization of the LMO balls for Muon and F-Muon for CNN training.}}
  {\includegraphics[width=0.6\linewidth]{figs/norms/fstardual_cifar.pdf}}
\end{figure}

@@

We evaluate F-Muon and S-Muon with $\alpha=0.5$ against Muon, SignSGD, and NSGD on the NanoGPT speedrun benchmark \citep{modded_nanogpt_2024}. The optimal hyperparameters are shown in the legend of \figureref{fig:nano_gpt_loss}, with \texttt{sign\_sgd\_coeff} values of 3e-4/0.07 for S-Muon. After 1750 training steps, F-Muon achieves a cross-entropy loss of 3.281, S-Muon achieves 3.287, and Muon achieves 3.279, falling below the target threshold of 3.280. As illustrated in \figureref{fig:nano_gpt_loss}, these differences are negligible. Notably, F-Muon represents a convex combination of Muon and NSGD, despite NSGD performing poorly in isolation. The same observation holds for S-Muon.

\begin{figure}[htbp]
\floatconts
  {fig:nano_gpt_loss}
  {\caption{The validation loss for NanoGPT}}
  {\includegraphics[width=0.7\linewidth]{figs/nanogpt/nano/val_loss_vs_step.pdf}}
\end{figure}

As observed in the CIFAR airbench experiments, setting $\alpha=1.1$ for F-Muon (corresponding to the no-ball configuration) yields a loss of 3.2818, representing only a marginal difference from the baseline.

Notably, F-Muon with the NSGD component not scaled by Muon's rule yields a loss of 3.288 (an increase of 0.007), while S-Muon scaled with this rule achieves a loss of 3.292 (an increase of 0.005).

@@

\begin{figure}[htbp]
  \floatconts
    {fig:med_gpt_loss}
    {\caption{The validation loss for GPT-2 Medium}}
    {\includegraphics[width=0.7\linewidth]{figs/nanogpt/medium/val_loss_vs_step.pdf}}
  \end{figure}

@@

The updates of LMO optimizers exhibit striking similarities between vector $l_p$ norms and matrix Schatten $S_p$ norms, as illustrated in \tableref{tbl:mat_vs_vec_lmo}. These parallels extend beyond the update formulas themselves to empirical performance characteristics. SignSGD closely resembles Adam, as noted in \citep{bernstein2024oldoptimizernewnorm}, and both Adam and Muon perform well during training large models \citep{zhao2025deconstructing, liu2025muon}. NSGD maintains identical updates in both vector and matrix cases. Greedy Coordinate Descent is rarely applied to high-dimensional problems, which provides perspective on why one-rank Neon underperforms in such settings.

@@

A common limitation of these analyses is their focus on convergence measured by the gradient norm. As we showed in our CIFAR experiments \sectionref{sec:cifar_underhood}, the gradient norm may decrease by only a factor of ten when the accuracy reaches 100\%.

@@

We generalized several successful algorithms, including Muon, Dion, and $\nu$SAM, to LMO-based algorithms using the family of norms dual to Ky Fan $k$-norms, yielding the Fanion family with low-rank updates. We further proposed the technique of regularizing these updates by combining them with NSGD or SignSGD, creating the F-Fanion and S-Fanion families through the $\norm{\cdot}_{\mathrm{F-KF-k}}^\dagger$ and $\norm{\cdot}_{\mathrm{C^\dagger-KF-k}}^\dagger$ norms. Our experiments demonstrate that F-Muon and S-Muon achieve competitive performance with Muon on CIFAR-10 airbench and NanoGPT tasks, confirming that the underlying norm constraint can be significantly modified without sacrificing effectiveness.

@@

\section*{Author Contributions}

IO suggested using the nuclear norm in the \citep{bernstein2024oldoptimizernewnorm} framework. DM presented the problem at the MIPT optimization course. IK suggested using composite norms (though not the ones that induce Fanions, F-Fanions or S-Fanions) and helped to draft and rewise the manuscript. NK suggested Lanczos algorithm as the most precise means to compute Fanions' updates, conducted experiments to prove this, and wrote \sectionref{sec:matrix-side}. AV conducted the finetuning of NanoGPT on Tiny Stories and wrote \sectionref{subsec:finetune}. All other work was done by AK: constructing Fanions, F-Fanions, S-Fanions, experimenting on the Linear Least Squares and CIFAR-airbench, pretraining NanoGPT and GPT-2 Medium, and writing the manuscript.

\section*{Acknowledgements}
We thank Dmitry Kovalev and Egor Shulgin for helpful discussions.

@@

\section{Proof of Lemma on the Dual to Convex Combinations}\label{sec:proof_dual_conv_comb}

@@

\begin{figure}[htbp]
\floatconts
  {fig:app_lls}
  {\caption{More images for Linear least squares problem for a 500x500 matrix}}
  {%
    \subfigure[The spectral norm of the gradient]{\label{fig:lls_op_grad_norm}%
      \includegraphics[width=0.5\linewidth]{figs/simple_lls/gradient_spectral_norm_vs_iteration_long_run.pdf}}%
    \vspace{1em}
    \subfigure[The nuclear norm of the gradient]{\label{fig:lls_nuclear_grad_norm}%
      \includegraphics[width=0.5\linewidth]{figs/simple_lls/gradient_nuclear_norm_vs_iteration_long_run.pdf}}%
    \vspace{1em}
    \subfigure[The loss over time]{\label{fig:lls_loss_time}%
      \includegraphics[width=0.5\linewidth]{figs/simple_lls/loss_vs_time_long_run.pdf}}
  }
\end{figure}

@@

Most bounds for Muon and other LMO-algorithms are given for the norm of the gradient \citep{li2025noteconvergencemuon, kovalev2025understanding, pethick2025training, riabinin2025gluon,kovalev2025noneuclideansgdstructuredoptimization}. Accordingly, it is natural to measure those norms on a real deep learning problem.

@@

We compare Muon, F-Muon, S-Muon, NSGD, SignSGD, Neon, F-Neon, Fanion-5, and F-Fanion-5 on CIFAR-airbench. NSGD and SignSGD are not heavily tuned. The hyperparameters of Neon, F-Neon, Fanion-5, and F-Fanion-5 are taken from Muon and F-Muon (see \tableref{tbl:cifar_lrs}). In the table, val\_accuracy is for the the weighted variant. However, to preserve the algorithms, we do not normalize the weights of the network each iteration.
